PCA

### 2023-11-11 DRAFT_PCA.Rmd
fixed more references and links for PCA page. full draft ready for review

Principal component analysis (PCA) is a multivariate method often used for summarising and visualising genetic data. In MarineOmics, PCA is commonly used to explore [population structure](/lnk to population structure section) and to highlight data issues such as batch effects, outliers, or variation to be explored (i.e. structural variation).

@@

Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) . It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see  for the OG description). Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see REF).  In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.

@@

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 1. 'Screeplot' of PCA eigenvalues, where there is one eigenvalue for every eigenvector. From Spies et al. 2022 (Supp. Fig. 2)."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 2. PCA plots from two studies (A) from Figure 2 in x, showing population structure results for 19 Ostrea lurida populations Plots of individual admixture determined using the program STRUCTURE (top) and PCA plots for PCs 1–5. PC1 is plotted against the latitude of the sampling site, then PC2 versus PC3 and PC4 versus PC5 and (B) from Figure 2 in x, showing PCA scores of 734 individuals of Atlantic Halibut (Hippoglossus hippoglossus) from samling sites on the North West Atlantic on PC1 and PC3 axes from PCA on 86 043 SNPs, highlighting population structure (PC3) and a putative inversion (PC1). In both plots, colors refer to the phylogeographic regions of each sample."}

@@

* Jolliffe, I. (2005). Principal component analysis. Wiley Online Library.
https://doi.org/10.1098/rsta.2015.0202
* Pearson K (1901) On lines and planes of closest fit to systems of points in space. Philos Mag
2:559–572
* Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377
* // add good reviews here: task Katherine Silliman and others
* Fentaw Abegaz et al.,2019. Principals about principal components in statistical genetics, Briefings in Bioinformatics. DOI: https://doi.org/10.1093/bib/bby08
* Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.

@@

This study by [Gompert & Buerkle, 2021](https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380) simulated a metapopulation along a 1-D stepping stone model with 50 patches (patch - a location in space), where dispersal was allowed only between adjacent patches, leading to isolation by distance (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380).

@@

D) 50 individuals were sampled from patches 20–31. Dark red and dark blue are used to denote peripheral patches in the sample and gray is used to denote central patches. In this case, sampling only the central patches also exaggerated the level of population structure. This horseshoe pattern in a PCA is a consequence of distance metrics that saturate (https://journals.asm.org/doi/10.1128/mSystems.00166-16). This saturation property arises in the case of  isolation by distance. With increasing distance between sampled patches, there is a loss of information about dissimilarity among patches (i.e. dissimilarity saturates with distance), and PCA cannot discriminate between samples that do not share any common features.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 3. Figure 4 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380"}

@@

**Number of SNPs**
The number of SNPs sampled has a large impact on the resolution of populations in PC space. In PCA, there is a critical signal-to-noise ratio below which the true structure of the signal cannot be recovered - in other words for genotype data, as genetic distance (i.e. FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases. The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see  McVean //add ref):

@@

### Batch effects {#batcheffects}
Batch effects are caused by technical differences among batches (i.e. groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space. Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). Batch effects can be removed with careful read trimming and filtering (see https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 and references therein).

As an example, let us consider simulated results from Lou & Therkildsen (2022; https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559) depicting batch effects related to differences in sequencing depth in a low-coverage whole genome dataset.  
Simulations consisted of nine populations on a 3x3 grid connected to neighbors in a low migration scenario (see Supporting Information at https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 for details). In this case, the PCA based on “true genotypes” should look similar to the figure below in which each of the populations are able to be distinguished as separate clusters in PC space.

```{r, echo = FALSE, include = TRUE, out.width = "30%", fig.align = "center", fig.cap="Figure 6. Example figure above modified from Figure 4 in https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077. This figure depicts patterns of spatial population structure inferred through principal component analysis (PCA) with lcWGS data. At top is a schematic for a scenario with nine populations (each coloured node corresponds to a simulated population) on a 3x3 grid connected to neighbors via gene flow, and below is the PCA based on the true genotypes under low migration conditions."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "100%", fig.align = "center", fig.cap="Figure 7. Differences in the sensitivity of three PCA programs to batch effects caused by different sequencing depths and number of samples. The number at the top of each panel is the number of individuals sampled from each simulated population. Color represents sequencing depth: yellow points mark individuals that were simulated sequenced at 0.125×; grey points for 4×. Figure 6 from https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559."}

@@

The following Figure 3 from Lotterhos 2019 (https://academic.oup.com/g3journal/article/9/6/1851/6028019) illustrates what happens in a PCA on landscape genomic data when there is recombination variation. She simulated a 2-D continuous landscape with local adaptation to an environmental cline, with a genome that consisted of linkage groups with quantitative trait loci that contributed to adaptation, a large neutral inversion (not involved in adaptation), and a region of low recombination (also not involved in adaptation). Even though the SNPs in the inversion made up less than 5% of the total number of SNPs in the PCA, and the inversion was not related to population structure, the first PC axis separated samples by inversion genotype (Figure A below - the second PC axes separated samples based on their haplotype in the region of low recombination).

When SNPs were thinned for LD and then a PCA was run on the data, the PCA showed a pattern of isolation-by-environment, which more accurately captured the population structure in the simulation (Figure B below - https://academic.oup.com/g3journal/article/9/6/1851/6028019).

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 8. Figure 3 from https://academic.oup.com/g3journal/article/9/6/1851/6028019."}

@@

Similar patterns have been observed for sex linked-markers. For example, https://onlinelibrary.wiley.com/doi/10.1111/mec.14217  found genetic clustering in a PCA by sex instead of by population structure, which was driven by as few as 12 sex-linked markers in the data.  Removing the sex-linked markers led to nonsignificant genetic structure in one species and a more accurate estimation of FST in the second species.

@@

Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) impute missing values. Imputation is the process of replacing a missing value with a numerical value via inference. Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see Yi & Latch, 2021). However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). Yi and Latch (2022) demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9. Figure 2(b) from Yi and Latch (2022) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares). Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right). As the % of missing data increases, the samples with the most missing data become centered at the PCA origin."}

@@

Imputation via the mean genotype within a site, sampling location, or population. 
In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population. The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 12). This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values. This method is not recommended (see Yi & Latch 2021), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10. Figure 4 from Yi and Latch (2022) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population. The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}

@@

### Hybridization and inbreeding effects {#hybrid}
Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ). However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding. These PCA patterns can be informative during early data exploration for identifying cryptic species or hybrids that you may want to remove prior to analyzing for population structure.

Below is Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380, with two parental species shown in dark red or dark blue, the F1 hybrids in yellow circles, and the F1-parental backcrosses in the lighter colors. Subsequent crosses between hybrids are shown in yellow with different symbols (F2 squares, F5 upward triangles, F20 downward triangles). F1...FN are genetically intermediate on PC1, and across all hybrids, PC1 mirrors the admixture proportion. In general, PC2 is associated with genetic variation among Fn generations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 11. Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380."}

@@

### Horseshoe and Wave patterns
Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them. The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from Morton et al below). In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis. Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (https://journals.asm.org/doi/10.1128/mSystems.00166-16). The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features. Figure 1A below shows an isolation by distance or “band” pattern, in which neighboring samples have mostly similar alleles, but the similarity in allelic composition declines with the distance between samples. For example, “Sample 0” has the reference alleles at SNPs 1-10, but does not share the same alleles at any SNPs with samples 10-20. As a result, the euclidean distance (on which a PCA is based) saturates with increasing distance between samples (Figure 1C below).

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 12. Figure 1 from https://journals.asm.org/doi/10.1128/mSystems.00166-16.  (a) A band table for haploids where SNPs are in rows, samples are in columns, and the color is the allele (0 or 1). Blocks that are colored black have a value of 1 while blocks that are colored white have a value of 0. (b) The first 2 components from a PCA of the band table, yielding the typical horseshoe shape. (c) The Euclidean distance (e.g., genetic relatedness) from point 0 to all of the other points. (d) An illustration of the distance saturation property."}

@@

The saturation property of PCA can also create “wave” patterns in a PCA of genetic data, which were initially misinterpreted to be signatures of specific migration events before they were shown to be statistical artifacts (Novembre & Stephens). Figure 2 from Novembre & Stephens below shows the pattern arising from simulating a 1-D stepping stone model, which is analogous to marine samples that are collected along a coastline. The horseshoe pattern arises in the first 2PC axes (Figure 2C), but this example also shows the wave patterns that arise on subsequent PC axes.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 13. Figure 2 from Novembre & Stephens. (a) Schematic of the one-dimensional habitat, with circles marking sampling locations and shades of blue marking order along the line. (b) One-dimensional PC maps (that is, plots of each PC element against the geographic position of the corresponding sample location). (c) Biplots of PC1 versus PC2, PC2 versus PC3, and PC3 versus PC4. Colors correspond to those in a. In many datasets without spatially referenced samples, the colors and the lines connecting neighboring points would not be observed; here they are shown to aid interpretation."}

@@

In MacVean (2009) Figure 5A below, PCA is carried out only on the haplotypes from CEU and YRI and all samples are subsequently projected onto the first PC identified from this analysis. Figure 5A shows how the loadings of individuals map onto PC1 each chromosome (in rows), and note the uniformity of the loadings for the source populations. However for the admixed group ASW, note how there is considerable variation at the level of individual chromosomes, with some chromosomes within an individual appearing essentially European (when a green point maps at the blue end) and others essentially African (when a green point maps at the orange end). 
Figure 5B below shows how the genome-wide admixture proportions can be inferred directly from the location of admixed samples on the first PC between the two source populations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 14. Figure 5 from https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000686. See text above for explanation."}

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:----------|:------------------------------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in (Dray & Josse, 2015); this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |@wold1975path|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
 |@purcell2007plink|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|

>>>>

Principal component analysis (PCA) is a multivariate method often used for summarising and visualising genetic data. In MarineOmics, PCA is commonly used to explore population structure and to highlight data issues such as batch effects, outliers, or variation to be explored (i.e. structural variation).

@@

Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) . It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see @Pearson1901-xe). Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see @Thioulouse2018-bj).  In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.

@@

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 1. 'Screeplot' of PCA eigenvalues, where there is one eigenvalue for every eigenvector. From @Spies2022-xj (Supp. Fig. 2)."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 2. PCA plots from two studies (A) from Figure 2 in @sillimanPopulationStructureGenetic2019, showing population structure results for 19 Ostrea lurida populations Plots of individual admixture determined using the program STRUCTURE (top) and PCA plots for PCs 1–5. PC1 is plotted against the latitude of the sampling site, then PC2 versus PC3 and PC4 versus PC5 and (B) from Figure 2 in @kessPutativeStructuralVariant2021, showing PCA scores of 734 individuals of Atlantic Halibut (Hippoglossus hippoglossus) from samling sites on the North West Atlantic on PC1 and PC3 axes from PCA on 86 043 SNPs, highlighting population structure (PC3) and a putative inversion (PC1). In both plots, colors refer to the phylogeographic regions of each sample."}

@@

* [Jolliffe, I. & Cadima, J. (2016) Principal component analysis: a review and recent developments](https://doi.org/10.1098/rsta.2015.0202)
* [Pearson K. (1901) On lines and planes of closest fit to systems of points in space. Philos Mag 2:559–572](https://doi.org/10.1080/14786440109462720)
* [Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377](https://doi.org/10.1007/978-1-4612-4380-9_14)
* [Fentaw Abegaz et al. (2019) Principals about principal components in statistical genetics, Briefings in Bioinformatics](https://doi.org/10.1093/bib/bby08)
* [Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.](http://www.numdam.org/item/JSFS_2012__153_2_79_0/)

@@

This study by @Gompert2016-wa simulated a metapopulation along a 1-D stepping stone model with 50 patches (patch - a location in space), where dispersal was allowed only between adjacent patches, leading to isolation by distance (@Gompert2016-wa).

@@

D) 50 individuals were sampled from patches 20–31. Dark red and dark blue are used to denote peripheral patches in the sample and gray is used to denote central patches. In this case, sampling only the central patches also exaggerated the level of population structure. This horseshoe pattern in a PCA is a consequence of distance metrics that saturate (@Morton2017-xy). This saturation property arises in the case of  isolation by distance. With increasing distance between sampled patches, there is a loss of information about dissimilarity among patches (i.e. dissimilarity saturates with distance), and PCA cannot discriminate between samples that do not share any common features.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 3. Figure 4 from @Gompert2016-wa"}

@@

**Number of SNPs**
The number of SNPs sampled has a large impact on the resolution of populations in PC space. In PCA, there is a critical signal-to-noise ratio below which the true structure of the signal cannot be recovered - in other words for genotype data, as genetic distance (i.e. FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases. The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms):

@@

### Batch effects {#batcheffects}
Batch effects are caused by technical differences among batches (i.e. groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (@Lou2022-lc). These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space. Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (@Lou2022-lc). Batch effects can be removed with careful read trimming and filtering (see @Lou2022-lc and references therein).

As an example, let us consider simulated results from @Lou2022-lc depicting batch effects related to differences in sequencing depth in a low-coverage whole genome dataset.  
Simulations consisted of nine populations on a 3x3 grid connected to neighbors in a low migration scenario (see Supporting Information at @Lou2022-lc for details). In this case, the PCA based on “true genotypes” should look similar to the figure below in which each of the populations are able to be distinguished as separate clusters in PC space.

```{r, echo = FALSE, include = TRUE, out.width = "30%", fig.align = "center", fig.cap="Figure 6. Example figure above modified from Figure 4 in @Lou2021-at. This figure depicts patterns of spatial population structure inferred through principal component analysis (PCA) with lcWGS data. At top is a schematic for a scenario with nine populations (each coloured node corresponds to a simulated population) on a 3x3 grid connected to neighbors via gene flow, and below is the PCA based on the true genotypes under low migration conditions."}

@@

```{r, echo = FALSE, include = TRUE, out.width = "100%", fig.align = "center", fig.cap="Figure 7. Differences in the sensitivity of three PCA programs to batch effects caused by different sequencing depths and number of samples. The number at the top of each panel is the number of individuals sampled from each simulated population. Color represents sequencing depth: yellow points mark individuals that were simulated sequenced at 0.125×; grey points for 4×. Figure 6 from @Lou2022-lc."}

@@

The following Figure 3 from Lotterhos 2019 (@Lotterhos2019-yz) illustrates what happens in a PCA on landscape genomic data when there is recombination variation. She simulated a 2-D continuous landscape with local adaptation to an environmental cline, with a genome that consisted of linkage groups with quantitative trait loci that contributed to adaptation, a large neutral inversion (not involved in adaptation), and a region of low recombination (also not involved in adaptation). Even though the SNPs in the inversion made up less than 5% of the total number of SNPs in the PCA, and the inversion was not related to population structure, the first PC axis separated samples by inversion genotype (Figure A below - the second PC axes separated samples based on their haplotype in the region of low recombination).

When SNPs were thinned for LD and then a PCA was run on the data, the PCA showed a pattern of isolation-by-environment, which more accurately captured the population structure in the simulation (Figure B below @Lotterhos2019-yz).

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 8. Figure 3 from @Lotterhos2019-yz."}

@@

Similar patterns have been observed for sex linked-markers. For example, @Benestan2017-oi found genetic clustering in a PCA by sex instead of by population structure, which was driven by as few as 12 sex-linked markers in the data.  Removing the sex-linked markers led to nonsignificant genetic structure in one species and a more accurate estimation of FST in the second species.

@@

Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) impute missing values. Imputation is the process of replacing a missing value with a numerical value via inference. Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see @Yi2022-bi). However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). @Yi2022-bi demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9. Figure 2(b) from @Yi2022-bi shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares). Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right). As the % of missing data increases, the samples with the most missing data become centered at the PCA origin."}

@@

Imputation via the mean genotype within a site, sampling location, or population. 
In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population. The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 12). This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values. This method is not recommended (see @Yi2022-bi), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10. Figure 4 from @Yi2022-bi shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population. The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}

@@

### Hybridization and inbreeding effects {#hybrid}
Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (@Gompert2016-wa). However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding. These PCA patterns can be informative during early data exploration for identifying cryptic species or hybrids that you may want to remove prior to analyzing for population structure.

Below is Figure 5 from @Gompert2016-wa, with two parental species shown in dark red or dark blue, the F1 hybrids in yellow circles, and the F1-parental backcrosses in the lighter colors. Subsequent crosses between hybrids are shown in yellow with different symbols (F2 squares, F5 upward triangles, F20 downward triangles). F1...FN are genetically intermediate on PC1, and across all hybrids, PC1 mirrors the admixture proportion. In general, PC2 is associated with genetic variation among Fn generations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 11. Figure 5 from @Gompert2016-wa."}

@@

### Horseshoe and Wave patterns
Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them. The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from @Morton2017-xy below). In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis. Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (@Morton2017-xy). The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features. Figure 1A below shows an isolation by distance or “band” pattern, in which neighboring samples have mostly similar alleles, but the similarity in allelic composition declines with the distance between samples. For example, “Sample 0” has the reference alleles at SNPs 1-10, but does not share the same alleles at any SNPs with samples 10-20. As a result, the euclidean distance (on which a PCA is based) saturates with increasing distance between samples (Figure 1C below).

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 12. Figure 1 from @Morton2017-xy.  (a) A band table for haploids where SNPs are in rows, samples are in columns, and the color is the allele (0 or 1). Blocks that are colored black have a value of 1 while blocks that are colored white have a value of 0. (b) The first 2 components from a PCA of the band table, yielding the typical horseshoe shape. (c) The Euclidean distance (e.g., genetic relatedness) from point 0 to all of the other points. (d) An illustration of the distance saturation property."}

@@

The saturation property of PCA can also create “wave” patterns in a PCA of genetic data, which were initially misinterpreted to be signatures of specific migration events before they were shown to be statistical artifacts (@Novembre2008-gs). Figure 2 from @Novembre2008-gs below shows the pattern arising from simulating a 1-D stepping stone model, which is analogous to marine samples that are collected along a coastline. The horseshoe pattern arises in the first 2PC axes (Figure 2C), but this example also shows the wave patterns that arise on subsequent PC axes.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 13. Figure 2 from @Novembre2008-gs. (a) Schematic of the one-dimensional habitat, with circles marking sampling locations and shades of blue marking order along the line. (b) One-dimensional PC maps (that is, plots of each PC element against the geographic position of the corresponding sample location). (c) Biplots of PC1 versus PC2, PC2 versus PC3, and PC3 versus PC4. Colors correspond to those in a. In many datasets without spatially referenced samples, the colors and the lines connecting neighboring points would not be observed; here they are shown to aid interpretation."}

@@

In @McVean2009-ms Figure 5A below, PCA is carried out only on the haplotypes from CEU and YRI and all samples are subsequently projected onto the first PC identified from this analysis. Figure 5A shows how the loadings of individuals map onto PC1 each chromosome (in rows), and note the uniformity of the loadings for the source populations. However for the admixed group ASW, note how there is considerable variation at the level of individual chromosomes, with some chromosomes within an individual appearing essentially European (when a green point maps at the blue end) and others essentially African (when a green point maps at the orange end). 
Figure 5B below shows how the genome-wide admixture proportions can be inferred directly from the location of admixed samples on the first PC between the two source populations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 14. Figure 5 from @McVean2009-ms. See text above for explanation."}

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:----------|:------------------------------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |@wold1975path|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
 |@purcell2007plink|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|
